Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-20 src/docs_paper/paper.Rmd
Add redundancy to HIV outputs plot using captions #55
Import point estimates for outputs plot #55

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
Two alternatives for the matrix decomposition [@jackel2005note] are
1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
2. the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
This estimate may be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

1. TMB (`r signif(time_taken$TMB, 2)` seconds), based on a Gaussian approximation at $\hat{\btheta}_\texttt{LA}$.
2. PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), based on a Gaussian approximation mixture at the nodes of a PCA-AGHQ quadrature grid, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
3. NUTS (`r signif(time_taken$tmbstan, 2)` days), the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling using Stan [@carpenter2017stan] implemented via the `tmbstan` package [@monnahan2018no].

@@

```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and HIV incidence and new cases (C) for adults 15-49 in Malawi. Inference conducted using TMB."}
knitr::include_graphics("figB.png")
```

@@

## Inference comparison \label{sec:inf-comparison}

To compare the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS we assessed (1) marginal point estimates, (2) marginal Kolmogorov-Smirnov and Anderson-Darling tests using the empirical cumulative distribution function (ECDF), (3) joint Pareto-smoothed importance sampling results, and (4) joint maximum mean discrepancy results.

@@

The root mean square error (RMSE) between posterior mean estimates from PCA-AGHQ and NUTS (`r rmse_ahgq_mean`) was `r abs(rmse_diff_mean)`% lower than that between TMB and NUTS (`r rmse_tmb_mean`).
For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_ahgq_sd)` (PCA-AGHQ).
These results, alongside those for the mean absolute error (MAE), are presented in Figure \ref{fig:mean-sd}.

```{r mean-sd, fig.cap="PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
knitr::include_graphics("depends/mean-sd-alt-latent.png")
```

>>>>

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
Two alternatives for the matrix decomposition [@jackel2005note] are
(1) the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
(2) the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
This estimate may be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

1. TMB (`r signif(time_taken$TMB, 2)` seconds), based on a Gaussian approximation at $\hat{\btheta}_\texttt{LA}$.
2. PCA-AGHQ (`r signif(time_taken$aghq, 2)` hours), based on a Gaussian approximation mixture at the adapted nodes $\z \in \mathcal{Q}(m, s, k)$, as described in Section \ref{sec:pca}, and implemented via extension of the `aghq` package [@stringer2021implementing].
3. NUTS (`r signif(time_taken$tmbstan, 2)` days), the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling using Stan [@carpenter2017stan] implemented via the `tmbstan` package [@monnahan2018no].

@@

```{r naomi-results, fig.cap="District-level HIV prevalence (A), ART coverage (B), and new HIV cases and HIV incidence (C) for adults 15-49 in Malawi. Inference conducted using TMB."}
knitr::include_graphics("figB.png")
```

@@

## Inference comparison\label{sec:inf-comparison}

We compared the accuracy of posterior distributions produced by TMB and PCA-AGHQ as compared with those from NUTS for latent field parameters and model outputs.
The metrics we used were (1) marginal point estimates, (2) marginal Kolmogorov-Smirnov and Anderson-Darling tests using the empirical cumulative distribution function (ECDF), (3) joint Pareto-smoothed importance sampling results, and (4) joint maximum mean discrepancy results.

@@

For the latent field, the root mean square error (RMSE) between posterior mean estimates from PCA-AGHQ and NUTS (`r rmse_ahgq_mean`) was `r abs(rmse_diff_mean)`% lower than that between TMB and NUTS (`r rmse_tmb_mean`).
For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_ahgq_sd)` (PCA-AGHQ).
These results, alongside those for the mean absolute error (MAE), are presented in Figure \ref{fig:mean-sd-latent}.

These improvements did not transfer to the model outputs (Figure \ref{fig:mean-sd-output}).

```{r mean-sd-latent, fig.cap="For the latent field PCA-AGHQ modestly improves estimation of the posterior mean, and substantially improves estimation of the posterior standard deviation, as compared with TMB."}
knitr::include_graphics("depends/mean-sd-alt-latent.png")
```

```{r mean-sd-output, fig.cap="PCA-AGHQ doesn't do so well for the model outputs."}
knitr::include_graphics("depends/mean-sd-alt-output.png")
```
